---
title: "COVID Flu: Flu Titer"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: True
editor_options:
  chunk_output_type: console
---

Tracking the titers of 4 flu strains pre- (D0) and post- (D28) flu vaccination
```{r setup, message=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(SummarizedExperiment)
library(emmeans)
library(BiocParallel)
param = SnowParam(32,"SOCK",progressbar = T)
register(param)

source('R/utils/dream.models.R')
loadLibraries()

knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE, out.width=900)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE, out.width=900)

output.folder <- file.path("output","titer")
output.prefix <- "flu.titer."
dir.create(output.folder)
figure.folder <- "figures"
dir.create(figure.folder)
```


#### Import and transform data
Convert titers outside of detection range into fixed values
```{r transform}
# Titer D0 and D28
titer.data <- read.csv(file.path("data","covid_flu.influenza.antibody.titer.and.avidity.data.csv"),check.names = F,na.strings = "")
titer.data <- titer.data[,1:6]
titer.params <- colnames(titer.data)[3:6]
titer.data <- titer.data[rowSums(is.na(titer.data[titer.params])) == 0,]

# Subject info from RNA-Seq data
rna.count.data <- readRDS(file.path("data","covid_flu.bulk.rna.count.data.RDS"))
subject.visit.info <- as.data.frame(colData(rna.count.data))
subject.visit.info <- unique(subject.visit.info[,1:10])
subject.visit.info$group <- factor(subject.visit.info$group,levels = c("HC","COVR"))

joined.data <- merge(subject.visit.info,titer.data,by=c("alt.subject.id","visit"))
# change all < numbers to 10 and > numbers to 2600
joined.data[,titer.params] <- apply(joined.data[,titer.params],2,function(x){x[grep("<",x)] <- "10";return(x)})
joined.data[,titer.params] <- apply(joined.data[,titer.params],2,function(x){x[grep(">",x)] <- "2600";return(x)})
joined.data[,titer.params] <- apply(joined.data[,titer.params],2,as.numeric)
rownames(joined.data) <- paste0(joined.data$alt.subject.id,".",gsub("Day ","D",joined.data$visit))

# max of scaled titer
joined.data$max.scaled.log2.titer <- apply(scale(log2(joined.data[,titer.params])),1,max)
joined.data$max.scaled.titer <- apply(scale(joined.data[,titer.params]),1,max)
```

#### D28/D0 Fold Change
Only include subjects under 65 years of age
```{r fc}
joined.data.under65 <- subset(joined.data,age < 65)
titer.fc <- calculate.fold.change(log2(t(joined.data.under65[,titer.params])),joined.data.under65,c("Day 0"))
titer.fc <- round(t(2^titer.fc$fc[,grep("D28$",colnames(titer.fc$fc))]),10) # numerical instability
titer.fc <- merge(joined.data[,1:10],titer.fc,by="row.names",suf=c("",".FC"))
write.csv(titer.fc,file.path(output.folder,paste0(output.prefix,"d28.fold.change.csv")),row.names = F)
```

#### Modeling
NB Model 1: FC ~ overall.group.sex + age + race + flu.vax.count.10yr + log2(D0.titer)
log-normal Model 2: D28.titer ~ overall.group.sex + age + race + flu.vax.count.10yr + log2(D0.titer)
based on FDA flu vaccine statistical analysis (e.g. page 27 of https://fda.gov/media/135687/download)
```{r model}
male.contrast <- "COVR.Male - HC.Male"
female.contrast <- "COVR.Female - HC.Female"
cat("Negative binomial model:\n")
fc.model.res <- data.frame()
model.data <- merge(subset(joined.data.under65,visit == "Day 0"),titer.fc[,c("alt.subject.id",titer.params)],by="alt.subject.id",suf=c(".D0",".FC"))
model.data$overall.group.sex <- paste0(model.data$group,".",model.data$sex)
for (i in titer.params) {
  model.data$D0.titer <- model.data[,paste0(i,".D0")]
  model.data$FC <- model.data[,paste0(i,".FC")]
  lm.res <- MASS::glm.nb(FC ~ overall.group.sex + age + race + flu.vax.count.10yr + log(D0.titer),model.data)
  em <- emmeans(lm.res, ~ overall.group.sex)
  fc.model.res <- rbind(fc.model.res,cbind(strain = i,as.data.frame(pairs(em,adjust="none"))))
}
cat("Contrast:",male.contrast,"\n")
subset(fc.model.res,contrast == male.contrast)

cat("Contrast:",female.contrast,"\n")
subset(fc.model.res,contrast == female.contrast)
write.csv(subset(fc.model.res,contrast %in% c(male.contrast,female.contrast)),
          file.path(output.folder,paste0(output.prefix,"FC.model.res.csv")))

cat("log-normal model:\n")
baseline.titer.model.res <- data.frame()
d28.titer.model.res <- data.frame()
model.data <- merge(subset(joined.data.under65,visit == "Day 0"),
                    subset(joined.data.under65,visit == "Day 28")[,c("alt.subject.id",titer.params,"max.scaled.log2.titer")],by="alt.subject.id",suf=c(".D0",".D28"))
model.data$overall.group.sex <- paste0(model.data$group,".",model.data$sex)
for (i in c(titer.params,"max.scaled.log2.titer")) {
  # baseline
  model.data$titer.D0 <- model.data[,paste0(i,".D0")]
  if (i == "max.scaled.log2.titer") {
    lm.res <- lm(titer.D0 ~ overall.group.sex + age + race,model.data)
  } else {
    lm.res <- lm(log10(titer.D0) ~ overall.group.sex + age + race,model.data)
  }
  em <- emmeans(lm.res, ~ overall.group.sex)
  baseline.titer.model.res <- rbind(baseline.titer.model.res,cbind(strain = i,as.data.frame(pairs(em,adjust="none"))))
  
  # day 28
  model.data$titer.D28 <- model.data[,paste0(i,".D28")]
  if (i == "max.scaled.log2.titer") {
    lm.res <- lm(titer.D28 ~ overall.group.sex + age + race + flu.vax.count.10yr + titer.D0,model.data)
  } else {
    lm.res <- lm(log10(titer.D28) ~ overall.group.sex + age + race + flu.vax.count.10yr + log10(titer.D0),model.data)
  }
  em <- emmeans(lm.res, ~ overall.group.sex)
  d28.titer.model.res <- rbind(d28.titer.model.res,cbind(strain = i,as.data.frame(pairs(em,adjust="none"))))
}
write.csv(subset(baseline.titer.model.res,contrast %in% c(male.contrast,female.contrast)),
          file.path(output.folder,paste0(output.prefix,"D0.model.res.csv")))

cat("Contrast:",male.contrast,"\n")
subset(d28.titer.model.res,contrast == male.contrast)

cat("Contrast:",female.contrast,"\n")
subset(d28.titer.model.res,contrast == female.contrast)
write.csv(subset(d28.titer.model.res,contrast %in% c(male.contrast,female.contrast)),
          file.path(output.folder,paste0(output.prefix,"D28.model.res.csv")))
```

#### FC Plots
```{r fc.plot,fig.cap=c("Fig. 2h","Fig. 2i")}
geo.mean <- function(x){exp(mean(log(x)))}
# geometric mean of fold change
gmfr <- titer.fc %>% group_by(group,sex) %>% summarise_at(.vars=titer.params,geo.mean)
# individual strain FC
titer.fc.melted <- reshape2::melt(titer.fc,measure.vars = titer.params,variable.name = "strain")

colors <- setNames(c(alpha(pal_nejm()(2),0.3),pal_jama()(2)),c("Female","Male","COVR","HC"))
plots <- list()
for (s in unique(titer.fc.melted$sex)) {
  model.stat <- subset(fc.model.res,contrast == paste0("COVR.",s," - HC.",s))
  model.stat$strain <- gsub(".","/",model.stat$strain,fixed=T)
  model.stat$group1 <- factor("HC",levels = c("HC","COVR"))
  model.stat$group2 <- factor("COVR",levels = c("HC","COVR"))
  model.stat$y.position <- 5.5
  model.stat$p.value <- round(model.stat$p.value,3)
  
  plots[[s]] <- ggplot(subset(titer.fc.melted,sex == s),aes(group,value)) + 
          geom_dotplot(binaxis = "y",stackdir="center",binwidth = 0.1,dotsize = 1.5,stackratio = 0.8,aes(fill=sex,color=NA)) +
          stat_summary(fun = mean, fun.min = mean, fun.max = mean, geom = "crossbar",width = 0.75,alpha=0.7,aes(color=group,fill=group)) +
          stat_pvalue_manual(data=model.stat,label = "p.value",xmin = "group1",xmax="group2",y.position = "y.position",size=3) +
          scale_y_continuous(trans = "log2",limits = c(NA,56))  +
          facet_grid(.~strain) + theme_pubr() + ylab("D28/D0 titer FC") + 
          scale_color_manual(values=colors) + scale_fill_manual(values=colors) + 
          scale_alpha_manual(values=c(0.4,0.9)) + 
          theme(strip.text = element_text(size=6,face = "bold"),
                strip.background = element_rect(fill=NA,color=NA),
                text = element_text(size=9),legend.position = "none") +
          ggtitle(s) + xlab("")
}
plots[["Female"]]
ggsave(file.path(figure.folder,"Fig2h.pdf"),width=5.25,height=2.5)
plots[["Male"]]
ggsave(file.path(figure.folder,"Fig2i.pdf"),width=5.25,height=2.5)

```

#### Titer Plots
```{r titer.plots,fig.cap=c("ED Fig. 3e","ED Fig. 3f","ED Fig. 3g")}
joined.data.under65.melted <- reshape2::melt(joined.data.under65[,c("alt.subject.id","visit","group","sex",titer.params,"max.scaled.log2.titer")],
                                             variable.name="strain")
# baseline titers
plots <- list()
for (s in unique(joined.data.under65.melted$sex)) {
  model.stat <- subset(baseline.titer.model.res,contrast == paste0("COVR.",s," - HC.",s) & strain != "max.scaled.log2.titer")
  model.stat$group1 <- factor("HC",levels = c("HC","COVR"))
  model.stat$group2 <- factor("COVR",levels = c("HC","COVR"))
  model.stat$y.position <- 3.8
  model.stat$p.value <- round(model.stat$p.value,3)
  
  plots[[s]] <- ggplot(subset(joined.data.under65.melted,visit == "Day 0" & sex == s & strain != "max.scaled.log2.titer"),aes(group,value)) + 
          geom_boxplot(outlier.shape = NA,width=0.8,aes(color=group)) + 
          geom_dotplot(binaxis = "y",stackdir="center",dotsize = 1,aes(color=sex,fill=sex)) +
          stat_pvalue_manual(data=model.stat,label = "p.value",xmin = "group1",xmax="group2",y.position = "y.position",size=3) +
          scale_y_continuous(trans = "log10",limits = c(NA,10000)) +           
          scale_fill_manual(values=alpha(colors,0.4)) + scale_color_manual(values=colors) +
          theme_classic() + ggtitle(s) + ylab("D0 MN Titers") + xlab("") + facet_grid(.~strain) + 
          theme(strip.text = element_text(size=9,face="bold"),strip.background = element_blank())
}
plots[["Female"]]
ggsave(file.path(figure.folder,"ED.Fig3e.pdf"),width=7,height=3)
plots[["Male"]]
ggsave(file.path(figure.folder,"ED.Fig3f.pdf"),width=7,height=3)

# D28 max titer
model.stat <- subset(d28.titer.model.res,contrast %in% c(male.contrast,female.contrast) & strain == "max.scaled.log2.titer")
model.stat$group1 <- factor("HC",levels = c("HC","COVR"))
model.stat$group2 <- factor("COVR",levels = c("HC","COVR"))
model.stat$sex <- ifelse(model.stat$contrast == male.contrast,"Male","Female")
model.stat$y.position <- 2.9
model.stat$p.value <- round(model.stat$p.value,3)
  
ggplot(subset(joined.data.under65.melted,visit == "Day 28" & strain == "max.scaled.log2.titer"),aes(group,value)) + 
    geom_boxplot(outlier.shape = NA,aes(color=group)) +
    geom_dotplot(binaxis = "y",stackdir="center",alpha=0.65,dotsize = 1,binwidth = 0.1,aes(fill=sex)) +
    stat_pvalue_manual(data=model.stat,label = "p.value",xmin = "group1",xmax="group2",y.position = "y.position",size=3) +
    scale_color_manual(values=colors) + scale_fill_manual(values=colors) + ylim(NA,3) + facet_grid(~sex) +
    theme_classic() + ylab("D28 max. standardized titers") + xlab("")
ggsave(file.path(figure.folder,"ED.Fig3g.pdf"),width=5,height=3)

```

#### Session Info
```{r session}
sessionInfo()
```