---
title: "COVID Flu: Plasmablast Frequencies"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: True
editor_options:
  chunk_output_type: console
---

```{r setup,message=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(SummarizedExperiment)
source('R/utils/dream.models.R')
loadLibraries()

knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE, out.width=900)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE, out.width=900)

output.folder <- file.path("output","flow.and.cbc")
dir.create(output.folder)
figure.folder <- "figures"
dir.create(figure.folder)
output.prefix <- "covid_flu.cytek.pb."

# cytek cell PB frequencies
cytek.data.pb <- read.csv(file.path("data","covid_flu.cytek.Bcell.panel.plasmablast.counts.csv"),check.names = F)

# Subject data within RNA-Seq data
normalized.rna.data <- readRDS(file.path("data","covid_flu.bulk.rna.count.data.RDS"))
sample.info <- as.data.frame(colData(normalized.rna.data))
subject.visit.info <- unique(sample.info[,1:9])

rownames(subject.visit.info) <- paste0(subject.visit.info$alt.subject.id,".",subject.visit.info$visit)
```

#### Normalization
Calculate b cell frequencies as % lymphocytes
```{r normalize.pb}
normalize.by <- "Lymphocytes"
cytek.data.pb[,-c(1:3)] <- cytek.data.pb[,-c(1:3)]/cytek.data.pb[,normalize.by]
# only PB populations are further analyzed
cytek.populations <- colnames(cytek.data.pb)
cytek.populations <- cytek.populations[-c(1:which(cytek.populations == "Live CD19+"))]
pb.flu.specific.populations <- grep("H3[+-]|HA",cytek.populations,value=T)

cytek.pb.freq <- merge(subject.visit.info,cytek.data.pb,by=c("alt.subject.id","visit"),all.y=T)
cat("Number of subjects:",length(unique(cytek.pb.freq$alt.subject.id)),"\n")
table(cytek.pb.freq$visit)

saveRDS(cytek.pb.freq,file.path(output.folder,paste0(output.prefix,"freq.as.prop.of.",normalize.by,".RDS")))

cytek.pb.freq.under65 <- subset(cytek.pb.freq,age < 65)
```

#### Baseline differential analysis
what are the differences at day 0 before flu vaccination?
```{r baseline.model}
baseline.cytek.pb.freq <- subset(cytek.pb.freq,visit == "Day 0")
freq.w.offset <- add.dynamic.offset(baseline.cytek.pb.freq[,cytek.populations])
baseline.model.fit <- baseline.condition.model(log2(t(freq.w.offset)),
                                               baseline.cytek.pb.freq,random.effects = "batch")
baseline.model.de.res <- model.term.statistics(baseline.model.fit,c("COVR.v.HC","COVR.male.v.female","COVR.m","COVR.f"))

write.csv(baseline.model.de.res,
          file.path(output.folder,paste0(output.prefix,"freq.as.prop.of.",normalize.by,".difference.at.baseline.csv")),row.names = F)
```

#### Association with time since diagnosis
Are any of the cell frequencies associated with TSD at baseline?
```{r tsd.asso.model}
baseline.cytek.pb.freq.w.tsd <- subset(baseline.cytek.pb.freq,!is.na(covid.diagnosis.start.date.to.sample.drawn))

tsd.model.fit <- baseline.tsd.association.model(t(apply(baseline.cytek.pb.freq.w.tsd[,cytek.populations],2,rank)),
                                               baseline.cytek.pb.freq.w.tsd,random.effects = "batch")
tsd.model.de.res <- model.term.statistics(tsd.model.fit,grep("TSD",colnames(tsd.model.fit$coefficients),value=T))
write.csv(tsd.model.de.res,file.path(output.folder,paste0(output.prefix,"freq.as.prop.of.",normalize.by,".tsd.association.at.baseline.csv")),row.names = F)

```

#### Contrast vaccination timepoints with baseline
Include only subjects < 65 years of age
```{r timepoint.model}
freq.w.offset <- add.dynamic.offset(cytek.pb.freq.under65[,cytek.populations])
timepoint.model.fit <- timepoint.condition.model(log2(t(freq.w.offset)),
                                                     cytek.pb.freq.under65,c("Day 0"),additional.fixed.effects = "batch")
timepoint.model.de.res <- model.term.statistics(timepoint.model.fit,grep("^(Baseline|D7)",colnames(timepoint.model.fit$coefficients),value = T))
timepoint.model.de.res[,c("visit","contrast")] <- do.call(rbind,sapply(timepoint.model.de.res$term,stringi::stri_split_fixed,pattern=".",n=2))
write.csv(timepoint.model.de.res,
          file.path(output.folder,paste0(output.prefix,"freq.as.prop.of.",normalize.by,".timepoint.diff.under65.csv")),row.names = F)
```

#### Visualization of PB Changes
Fig. 2g
```{r pb}
cytek.pb.freq.under65$max.flu.specific.pb <- cytek.pb.freq.under65$`All HA+(CD20lowCD21low PB)`
# recalculate as fraction of CD19+ B-cells
cytek.pb.freq.under65$max.flu.specific.pb <- cytek.pb.freq.under65$max.flu.specific.pb/cytek.pb.freq.under65$`CD19+`
cytek.pb.freq.under65$overall.group.visit <- factor(paste0(cytek.pb.freq.under65$group,"\n",gsub("Day ","D",cytek.pb.freq.under65$visit)),
                                                    levels=c("HC\nD0","HC\nD7","COVR\nD0","COVR\nD7"))
colors <- setNames(c(pal_nejm()(2),pal_jama()(2)),c("Female","Male","COVR","HC"))

ggplot(cytek.pb.freq.under65,aes(overall.group.visit,1000*max.flu.specific.pb+.1)) + 
  geom_boxplot(outlier.shape = NA,width=0.8,aes(color=group)) + 
  geom_point(aes(color=sex,fill=sex),pch=21,size=0.5) + geom_line(aes(group=alt.subject.id,color=sex),alpha=0.8) +
  stat_compare_means(comparisons = list(c(1,3),c(2,4)),label = "p.format",size=2.5) +
  facet_grid(~sex) + theme_pubr() + scale_color_manual(values=colors) + scale_fill_manual(values=alpha(colors,0.4)) +
  xlab("") + scale_y_continuous(trans="log10") + ylab(paste0("Flu-specific\nplasmablast frequency","\n(","per 1,000 B-cells",")")) +
  theme(strip.background = element_blank(),legend.position = "none")
ggsave(file.path(figure.folder,"Fig2g.pdf"),width=5.5,height=3)
```

####Session Info
```{r}
sessionInfo()
```
