---
title: "acute_covid_pbulk_monosig_score"
output: html_document
date: "2023-01-09"
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(edgeR)
library(GSVA)
library(tidyverse)
library(Biobase)
library(reshape2)
library(ggpubr)
library(DESeq2)
```

## Get GSVA score of the filtered pseudobulk objects for acute COVID data
acute COVID data from Liu et al, Cell, 2021  
using monocyte repressed BTM-M4.0/M11.0 innate immune receptor signature in Figure 1  
Mono baseline (D0, covid-healthy) BTM-M4.0/M11.0 innate immune receptor signature  
Input is output from filtered and normalized pseudobulk objects of acute COVID data GEO: GSE161918  
Download the Seurat object from GSE161918
And put it in the data/other_datasets/acute_covid/ folder
Generate pseudobulk lists as reported in https://github.com/niaid/covid19-time-resolved 
**Generate Extended Data Figure 1i boxplot**  
```{r input}
# Generate pseudobulk lists as reported in https://github.com/niaid/covid19-time-resolved
DGELISTS_IN_PATH <- c("data/other_datasets/acute_covid/pseudobulk_dgelists_normalized/Unsorted-WCTcoursecelltype.rds")
OUT_DIR <- "output/pbulk/gsva/"
dir.create(OUT_DIR, recursive = TRUE)

# input LE sets from dream model
selected.reset.M4.M11 <- readRDS("data/LE_sets/mono.LE.and.reset.genesets.RDS")
pbulk_list <- readRDS(DGELISTS_IN_PATH)

# list of expression data by cell type
eset_list <- lapply(pbulk_list, function(dge){
  mat <- DESeq2::varianceStabilizingTransformation(dge$counts)
  meta <- dge$samples
  ExpressionSet(assayData = mat, phenoData = AnnotatedDataFrame(meta))
})
eset_list <- unlist(eset_list)
# remove technical control CHI014
eset_list <- lapply(eset_list, function(dge){
  dge <- dge[, dge$Donor != "CHI014"]
  dge$sample_id <- paste(dge$Donor, dge$Timepoint, sep = "_")
  return(dge)
})
sapply(eset_list, ncol)

eset_list_mono <- eset_list[c("Mono_Classical", "Mono_NonClassical")]
```
  
Calculate module scores  
```{r cal_score}
scores_list <- lapply(1:length(eset_list_mono), function(eset){
  cat(names(eset_list_mono)[eset],"\n")
  celltype.gsea.res <- selected.reset.M4.M11[[names(eset_list_mono)[eset]]]$CITEseq.baseline.LE.Union.genesets
  # if (nrow(celltype.gsea.res) > 0) {
  module.scores <- gsva(expr = eset_list_mono[[eset]], gset.idx.list = celltype.gsea.res, method = "gsva", parallel.sz = 16, min.sz = 5)
  return(cbind(reshape2::melt(exprs(module.scores)),celltype= names(eset_list_mono)[eset]))
  # }
})

module.scores.df <- do.call("rbind",scores_list)
colnames(module.scores.df) <- c("geneset","sample","module.score","celltype")
# add meta data and transform the dataframe into esetlist
sample_meta <- pbulk_list$Mono_Classical$samples %>% rownames_to_column(var = "sample")
# add other info
module.scores.df$geneset_celltype <- paste0(module.scores.df$geneset,"--",module.scores.df$celltype)
module.scores.df <- left_join(module.scores.df,
                              sample_meta[,c("sample", "Donor", "Age", "sex", "ever_admitted_to_icu", "Class", "cond_group", "Timepoint", "PC1_cat", "onset_group", "severity.outcome2", "PC1_onset_group", "days_since_onset","PC1")], 
                              by=c("sample"))
write.csv(module.scores.df, file.path(OUT_DIR, "module_score_gsva_filtered_samples_genes_df.Mono.M4M11.Brescia.csv"))
```

  
Boxplot of module scores  
Using the limma model to calculate p values given there are multiple samples from individuals  
Showing the union LEs from male and female module score  
```{r boxplot, fig.width=8, fig.height=6}
# module.scores.df <- read.csv(file.path(OUT_DIR, "module_score_gsva_filtered_samples_genes_df.Mono.M4M11.Brescia.csv"), header = TRUE, row.names = 1, stringsAsFactors = FALSE)
# test to account for the subjects with multiple timepoints -- use the limma model
source("utils/limma_DE_functions_cell_freq.R")
module.scores.exprs <- module.scores.df %>%
  dplyr::mutate(PC1_cat2 = factor(replace(as.character(.$PC1_cat), .$Class == "HC", "HC"), levels = c("HC","PC1_low","PC1_high"))) %>%
  filter(!is.na(PC1_cat2)) %>%
  reshape2::dcast(geneset_celltype~sample, value.var = "module.score") %>%
  column_to_rownames("geneset_celltype")

module.scores.meta <- module.scores.df %>%
  dplyr::mutate(PC1_cat2 = factor(replace(as.character(.$PC1_cat), .$Class == "HC", "HC"), levels = c("HC","PC1_low","PC1_high"))) %>%
  filter(!is.na(PC1_cat2)) %>%
  dplyr::distinct(sample, .keep_all = TRUE) %>%
  column_to_rownames("sample")

module.scores.meta[,c("geneset","module.score","celltype")] <- NULL
eset <- ExpressionSet(assayData = as.matrix(module.scores.exprs), phenoData = AnnotatedDataFrame(module.scores.meta))
FORMULA <- "~0 + PC1_cat2"
FORMULA <- as.formula(FORMULA)
design <- model.matrix(FORMULA, module.scores.meta)

colnames(design) <- gsub("PC1_cat2", "", colnames(design))
rownames(design) <- module.scores.meta$sample

fit <- RunLimma(eset, design_matrix = design, add_random_effect = TRUE, 
                block_this_variable = "Donor",
                do_contrast_fit = FALSE, my_contrast_matrix = NULL)

# make contrasts
CONTRAST <- c("PC1_low - HC","PC1_high - HC","PC1_high - PC1_low")
CONTRAST_NAME <- c("PC1_lowvsHC", "PC1_highvsHC", "PC1_highvsPC1_low")
toptab.list = list()

contrast_mat <- makeContrasts(contrasts = CONTRAST, levels = colnames(design)) 
colnames(contrast_mat) <- CONTRAST_NAME
  
# contrast_fit
cfit <- contrasts.fit(fit, contrasts = contrast_mat)
cfit <- eBayes(cfit)

for (i in 1:length(CONTRAST)){
  toptab <- topTable(cfit, number = nrow(cfit), coef = CONTRAST_NAME[[i]]) %>%
    rownames_to_column(var = "gene")
  toptab.list[[i]] <- toptab
}

names(toptab.list) <- CONTRAST_NAME
# saveRDS(toptab.list, file.path(OUT_DIR, "module_score_gsva.Mono.M4M11.Brescia.limmapval.rds"))
# toptab.list <- readRDS(file.path(OUT_DIR, "module_score_gsva.Mono.M4M11.Brescia.limmapval.rds"))

# add limma pvalue to the plot
toptab.df <- dplyr::bind_rows(toptab.list, .id = "contrast") %>%
  separate(col = "gene", into = c("geneset","celltype"), sep = "--") %>%
  separate(col = "contrast", into = c("group1","group2"), sep = "vs")
module.scores.df <- module.scores.df %>% 
  dplyr::mutate(PC1_cat2 = factor(replace(as.character(.$PC1_cat), .$Class == "HC", "HC"), levels = c("HC","PC1_low","PC1_high"))) %>%
  filter(!is.na(PC1_cat2))

# pdf("figures/Bresciadata.module.score.Monoc.M4M11.PC1.EDFig1i.pdf", width = 20, height = 6)
celltypes <- "Mono_Classical"
for (i in celltypes){
  df <- filter(module.scores.df, celltype == i, geneset == "m.and.f.union")
  pval <- filter(toptab.df, celltype == i, geneset == "m.and.f.union") %>% 
    mutate(y.position = c(0.5,0.7,0.4)) %>%
    mutate(P.Value = ifelse(P.Value < 0.01, formatC(P.Value, format = "e", digits = 2), formatC(P.Value, format = "f", digits = 2)))
  p <- ggplot(df, aes(x = PC1_cat2, y = module.score)) +
    geom_boxplot(aes(fill=PC1_cat2, color= PC1_cat2), alpha = 0.2, width = 0.3, outlier.shape = NA)+
    geom_point(aes(fill=PC1_cat2), shape = 21, color = "white", size = 4, position = position_jitterdodge(dodge.width = 0.9, jitter.width = 0.2)) +
    scale_fill_manual(values = c("HC" = "#e5862d", "PC1_low" = "#B4C6A6", "PC1_high" = "#FFCE45"))+
    scale_color_manual(values = c("HC" = "#e5862d", "PC1_low" = "#B4C6A6", "PC1_high" = "#FFCE45"))+
    # facet_wrap(~geneset, scale = "free", ncol = 4) +
    stat_pvalue_manual(pval, label = "P.Value", tip.length = 0.01,size = 7)+
    theme_bw(base_size = 26)+
    theme(axis.text.x = element_text(angle = 30, hjust = 1)) + 
    ylim(c(min(df$module.score), max(df$module.score)+0.3))+
    ggtitle(i)
  print(p)
}

# dev.off()
```

```{r}
sI <- sessionInfo()
utils:::print.sessionInfo(sI[-c(10,11)])
```

















