---
title: "acute_covid_scmono_M4M11sig"
output: html_document
date: "2023-01-09"
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(Seurat)
library(tidyverse)
library(matrixStats)
library(parallelDist)
library(RColorBrewer)
library(reshape2)
library(viridis)
library(ggrastr)
```


## Monocyte repressed innate immune signature expression in acute COVID data
## acute COVID data from Liu et al, Cell, 2021  
Download the Seurat object from GSE161918
And put it in the data/other_datasets/acute_covid/ folder
**Generate Extended Data Figure 1i dot plot**  

```{r input_seurat}
Brescia_covid <- readRDS(file.path("data","other_datasets","acute_covid","GSE161918_AllBatches_SeuratObj.rds"))
Brescia_mono <- subset(Brescia_covid, subset = Sorted == "N" & coursecelltype %in% c("Mono") & celltypeQC == TRUE)

# show the same genes as in main figure
selected.genes.mc <- c("PTAFR","TLR8","NOD2","CLEC4E","TLR2", "TLR5",
                       "LILRB3","FCER1G","FCGRT","SLC7A7","CYBB","CFP","SIRPA",
                       "SLC31A2","SCPEP1","RNF130","RASSF4","CYP1B1")
# do a scTransform to remove batch
Brescia_mono <- SCTransform(Brescia_mono, assay = "RNA", method = "glmGamPoi", vars.to.regress = "Batch", verbose = FALSE)
DefaultAssay(Brescia_mono) <- "SCT"
```
  
  
Classical Monocytes  
showing the signature genes as example in Figure 1 classical monocytes  
Dotplots  
**Extended Data Fig 1i**  

```{r dotplots, fig.width = 7, fig.height = 3}
Brescia_mono$PC1_cat2 <- replace(as.character(Brescia_mono$PC1_cat), as.character(Brescia_mono$Class) == "HC", "HC")
Brescia_mono$PC1_cat2 <- factor(Brescia_mono$PC1_cat2, levels = c("PC1_high","PC1_low","HC"))
Brescia_monoc <- subset(Brescia_mono, subset = WCTcoursecelltype == "Mono_Classical" & PC1_cat2 %in% c("HC","PC1_high","PC1_low"))

# pdf(file = "figures/Brescia.monoc.baselineLE.dotplot.EDFig1i.pdf", width = 7, height = 3)
DotPlot(Brescia_monoc, assay = "RNA", features = selected.genes.mc, group.by = "PC1_cat2", dot.scale = 10, cols = c("gray95", "blue"))+
  RotatedAxis()+ggtitle("D0 (Baseline) LE genes")+theme(axis.text=element_text(size=12))+
  scale_size(range=c(0,10))+
  scale_y_discrete(labels=c("More_severe", "Less_severe", "HC"))
# dev.off()
```


```{r}
sI <- sessionInfo()
utils:::print.sessionInfo(sI[-c(10,11)])
```







