---
title: "scTCR_circos_CD8_clonality_D0"
output: html_document
date: "2023-01-12"
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(circlize)
library(tidyverse)
library(ggpubr)
library(RColorBrewer)
```

## single cell TCR clonality in COVR CD8 subsets  
Plot D0 (baseline) CD8 subsets TCR clonality circos plot by connecting identical clones in each subject
**Generate circos plot for Extended Data Figure 4g**  
Single cell TCR data downloaded from Zenodo and put in data/TCR/ folder as input  

```{r input}
tcr_combined <- read.csv(file.path("data", "TCR", "tenx_filtered_scTCR_anno.csv"))
tcr_combined_CD8 <- filter(tcr_combined, coarse.cell.type == "CD8") %>%
  mutate(cell.type.GPR56 = paste(cell.type, GPR56_level, sep = "-")) %>%
  mutate(cellsubset = replace(.$cell.type, .$cell.type %in% c("CD8_EM","CD8_TEMRA"), .$cell.type.GPR56[which(.$cell.type %in% c("CD8_EM","CD8_TEMRA"))])) %>%
  filter(cellsubset %in% c("CD8_CM", "CD8_EM-GPR56-", "CD8_EM-GPR56+", "CD8_Naive", "CD8_TEMRA-GPR56+", "CD8_TRM")) %>%
  mutate(sample_id = sub(pattern = "Day ", replacement = "D", paste(alt.subject.id, visit, sep = ".")))

tcr_combined_CD8_filtered <- filter(tcr_combined_CD8, 
                                    group == "COVR",
                                    visit == "Day 0", cellsubset %in% c("CD8_Naive","CD8_CM","CD8_EM-GPR56+","CD8_EM-GPR56-","CD8_TEMRA-GPR56+"))

# further down sample, otherwise link too dense
# ds to 600 cells for each cell subset
set.seed(124)
tcr_combined_CD8_ds <- tcr_combined_CD8_filtered %>% 
  group_by(cellsubset) %>% 
  sample_n(size = 600, replace = F) %>%
  mutate(cellsubset.sex = paste(cellsubset, sex, sep = ".")) %>%
  mutate(cellsubset.sex = factor(cellsubset.sex, levels = c("CD8_EM-GPR56+.Female", "CD8_EM-GPR56+.Male", "CD8_EM-GPR56-.Female", "CD8_EM-GPR56-.Male", "CD8_TEMRA-GPR56+.Female", "CD8_TEMRA-GPR56+.Male", "CD8_CM.Female","CD8_CM.Male", "CD8_Naive.Female","CD8_Naive.Male")))
```

  
Label clones and index them  
```{r clone}
tcr_combined_CD8_ds2 <- dplyr::arrange(tcr_combined_CD8_ds, cellsubset.sex, group) %>%
  dplyr::ungroup() %>%
  mutate(celltype_rowindex = rep(c(1:unique(table(.$cellsubset))),5)) %>%
  mutate(rowindex = 1:nrow(.))
table(tcr_combined_CD8_ds2$cellsubset)
tcr_combined_CD8_ds2$sex_rowindex <- unlist(sapply(as.numeric(table(tcr_combined_CD8_ds2$cellsubset.sex)), seq_len))

clone_id <- data.frame(table(tcr_combined_CD8_ds2$CTaa_tcr)) %>%
  mutate(clone_id = paste("C",1:nrow(.), sep = ""))
tcr_combined_CD8_ds2 <- left_join(tcr_combined_CD8_ds2, clone_id, by = c("CTaa_tcr" = "Var1"))

tcr_cells = tcr_combined_CD8_ds2[,c("alt.subject.id","visit","sample_id","Timepoint.sex.group","sex", "group", "cellsubset", "celltype_rowindex", "rowindex", "clone_id", "barcode.batch","sex_rowindex", "cellsubset.sex")]
tcr_cells$cellsubset.subject <- paste(tcr_cells$cellsubset, tcr_cells$sex, tcr_cells$alt.subject.id, sep = ".")

```


Circos plots for CD8 subsets at D0 (baseline)  
**Extended Data Fig 4g**  
```{r circos, fig.width=8.6, fig.height=8.6}
n_cells <- nrow(tcr_cells)
tcr_cells$xmin <- 0
tcr_cells$xmax = n_cells
ring_colors <- list(c(brewer.pal(n = 5,name = "Set3")), 
                    c(brewer.pal(n = 8,name = "Set1"),
                      brewer.pal(n = 8,name = "Set2"),
                      brewer.pal(n = 8,name = "Paired")),
                    c("#BD3B29","#0472B6"))
# brewer.pal(n = 12,name = "Paired")))
names(ring_colors[[1]]) <- unique(tcr_cells$cellsubset)
names(ring_colors[[2]]) <- unique(tcr_cells$alt.subject.id)
names(ring_colors[[3]]) <- c("Female", "Male")
ring_colors_celltype <- rep(ring_colors[[1]], each = 10/5)
ring_colors_subject <- rep(ring_colors[[2]],5)
# ring_colors_sex <- rep(c("#BD3B29","#0472B6","#BD3B29","#0472B6","#BD3B29","#0472B6","#BD3B29","#0472B6","#BD3B29","#0472B6"),
#                        c(12,12,12,12,12,12,12,12,12,12))
ring_colors_sex <- rep(c("#BD3B29","#0472B6"),
                       5)

# link identical TCR clone with lines
tcr_links <- select(tcr_cells, c("alt.subject.id", "cellsubset.subject", "clone_id", "barcode.batch", "cellsubset")) %>%
  mutate(sample.clone = paste(cellsubset.subject, clone_id, sep = "."))
# select only duplicated rows and merge with all other clones
tcr_links_dup <- tcr_links[duplicated(tcr_links$sample.clone) | duplicated(tcr_links$sample.clone, fromLast = TRUE), ] %>%
  left_join(tcr_links[,c("sample.clone","barcode.batch")], by = "sample.clone")
tcr_links_dup <- tcr_links_dup[!tcr_links_dup$barcode.batch.x == tcr_links_dup$barcode.batch.y,]

# pdf("figures/CD8_circos_subsets_D0_EDFig4g.pdf", width = 8.6, height = 8.6)
{par(mar=rep(0,4))
circlize::circos.clear()
circlize::circos.par(
  cell.padding=c(0,0,0,0), track.margin=c(0,0.15), start.degree = 90, gap.degree = 1)

circlize::circos.initialize(factors=tcr_cells$cellsubset.sex,
                            xlim=matrix(c(rep(0, 10), table(tcr_cells$cellsubset.sex)), ncol=2))


# plot sectors
circlize::circos.trackPlotRegion(
  ylim = c(0, 1), 
  track.height = 0.2,
  bg.border=F,
  panel.fun = function(x, y) {
    name = circlize::get.cell.meta.data("sector.index")
    i = circlize::get.cell.meta.data("sector.numeric.index")
    xlim = circlize::get.cell.meta.data("xlim")
    ylim = circlize::get.cell.meta.data("ylim")
    
    circlize::circos.rect(
      xleft=xlim[1], ybottom=ylim[1],
      xright=xlim[2], ytop=ylim[2]-0.5,
      col = ring_colors_sex[i],
      border = ring_colors_sex[i])
    
    circlize::circos.rect(
      xleft=xlim[1], ybottom=ylim[1]+0.5,
      xright=xlim[2], ytop=ylim[2],
      col = ring_colors_celltype[i],
      border = ring_colors_celltype[i])
    
  })    


for (k in 1:nrow(tcr_links_dup)) {
  # for(k in 1){
  # determine row of tcr_cells to use for coloring
  i <- match(tcr_links_dup$barcode.batch.x[k], tcr_cells$barcode.batch)
  j <- match(tcr_links_dup$barcode.batch.y[k], tcr_cells$barcode.batch)
  
  # draw links, colored by selected variable
  circlize::circos.link(
    sector.index1=tcr_cells$cellsubset.sex[i], point1=c(tcr_cells$sex_rowindex[i]),
    sector.index2=tcr_cells$cellsubset.sex[j], point2=c(tcr_cells$sex_rowindex[j]),
    col = "grey", rou1=0.65, rou2=0.65, h.ratio = 0.4, lwd = 0.2
  )
}

legend(x = "topleft", legend = names(c(ring_colors[[3]],ring_colors[[1]])), fill = c(ring_colors[[3]],ring_colors[[1]]), bty = "n", y.intersp = 1)

circos.clear()
}

# dev.off()
```


SessionInfo  
```{r}
sI <- sessionInfo()
utils:::print.sessionInfo(sI[-c(10,11)])
```


