---
title: "mono_pbulksig_all_timepoints"
output: html_document
date: "2023-01-05"
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(edgeR)
library(GSVA)
library(Biobase)
library(reshape2)
library(tidyverse)
library(ggpubr)
library(openxlsx)
library(SummarizedExperiment)
library(BiocParallel)
library(doParallel)
param = SnowParam(32,"SOCK",progressbar = F)
register(param)
registerDoParallel()

source('utils/dream.models.R')
loadLibraries()

```

## Get GSVA score of acute COVID related gene signatures for Extended Data Figure 5d-5e
using the genesets that are significanlty enriched during acute COVID in monocytes and see their change after flu vaccination
input is output from filtered and normalized pseudobulk objects  
Leading-edge genes from acute COVID data Liu et al, Cell, 2021: Antigen presentation sets, HALLMARK-Inflammatory response, HALLMARK-TNFa signaling via NF-kB; MS1 monocyte signature from Reyes et al,Sci. Transl. Med., 2021

**Generate Extended Data Fig 5d and 5e, Supplementary Figure 2e**  
Seurat object needs to be downloaded and put in data/ folder, pseudobulk lists generated using the pooling functions  

```{r input}
# plotting functions
source("utils/boxplot_mono_pbulk_scores.R")
# Generate pseudobulk lists and set parameters for pbulk input/output
DGELISTS_IN_PATH <- "output/CITE-seq/covid_flu.CITEseq.pseudobulk.gene.expr.filtered.RDS"
OUT_DIR <- "output/pbulk/gsva/"
dir.create(OUT_DIR, recursive = TRUE)

pbulk_list <- readRDS(DGELISTS_IN_PATH)

# read in geneset list from Acute COVID cohort
celltypes = c("Mono_Classical","Mono_NonClassical")
Brescia.LElist <- list()
for (i in celltypes) {
  Brescia_mono_covid <- read.table(paste("data/other_datasets/acute_covid/", i, "--model@healthy_vs_covid--coef@COVID-Healthy--fgsea.tsv", sep = ""), header = TRUE, sep = "\t")
  Brescia_mono_pc1 <- read.table(paste("data/other_datasets/acute_covid/", i, "--model@PC1--coef@PC1--fgsea.tsv", sep = ""), header = TRUE, sep = "\t")

  mono_covid_Agpre <- Brescia_mono_covid %>% filter(pathway %in% c("reactome_Antigen processing-Cross presentation","reactome_MHC class II antigen presentation","KEGG_ANTIGEN_PROCESSING_AND_PRESENTATION"))
  mono_covid_Agpre_LE <- unique(unlist(sapply(mono_covid_Agpre$leadingEdge, function(x) str_split(x, pattern = " "))))
  mono_covid_NFkB <- Brescia_mono_covid %>% filter(pathway %in% c("HALLMARK_TNFA_SIGNALING_VIA_NFKB"))
  mono_covid_NFkB_LE <- unique(unlist(sapply(mono_covid_NFkB$leadingEdge, function(x) str_split(x, pattern = " "))))
  mono_covid_inflam <- Brescia_mono_covid %>% filter(pathway %in% c("HALLMARK_INFLAMMATORY_RESPONSE"))
  mono_covid_inflam_LE <- unique(unlist(sapply(mono_covid_inflam$leadingEdge, function(x) str_split(x, pattern = " "))))
  
  ### MS1 signature from Reyes et al, Sci. Transl. Med. (2021)
  MS1_module <- c("S100A8","VCAN","VIM","LGALS1","MNDA","CTSD","TIMP1","NCF1", "S100A12","LILRA5","SELL","MCEMP1","CLU","VAMP5","CYP1B1","CD163","ACSL1","SOD2", "ALOX5AP","CTSA","VNN2","IL1R2","STXBP2","NAMPT","CALR","PLAC8","ANXA6","LYZ","RETN","NKG7")
  Brescia.LElist[[i]] <- list("mono_covid_Agpre_LE" = mono_covid_Agpre_LE, "mono_covid_inflam_LE" = mono_covid_inflam_LE, 
                              "mono_covid_NFkB_LE" = mono_covid_NFkB_LE, "MS1_module" = MS1_module)
}

# list of expression data by cell type
eset_list <- lapply(pbulk_list, function(dge){
  mat <- dge$normalizedExpr
  meta <- dge$samples
  ExpressionSet(assayData = mat, phenoData = AnnotatedDataFrame(meta))
})

eset_list_mono <- eset_list[c("Mono_Classical", "Mono_NonClassical")]
```
  
  
Calculate module scores
```{r cal_score}
scores_list <- lapply(1:length(eset_list_mono), function(eset){
  cat(names(eset_list_mono)[eset],"\n")
  celltype.gsea.res <- Brescia.LElist[[names(eset_list_mono)[eset]]]
  # if (nrow(celltype.gsea.res) > 0) {
  module.scores <- gsva(expr = eset_list_mono[[eset]], gset.idx.list = celltype.gsea.res, method = "gsva", parallel.sz = 16, min.sz = 5)
  return(cbind(reshape2::melt(exprs(module.scores)),celltype= names(eset_list_mono)[eset]))
  # }
})

module.scores.df <- do.call("rbind",scores_list)
colnames(module.scores.df) <- c("geneset","sample","module.score","celltype")
# add meta data and transform the dataframe into esetlist
meta <- pbulk_list$Mono$samples[,c("alt.subject.id", "age", "Batch","visit","sex","race","group","flu.vax.count.10yr","vaccine.side.effects","long.covid.symptoms", "visit.overall.group.sex")] 
meta$sample <- rownames(meta)
meta$sample_id <- paste(meta$alt.subject.id, meta$visit.overall.group.sex, sep = ".")

# add other info
module.scores.df$geneset_celltype <- paste0(module.scores.df$geneset,"--",module.scores.df$celltype)
module.scores.df <- left_join(module.scores.df, 
                              meta,
                              by=c("sample"))
write.csv(module.scores.df, file.path(OUT_DIR, "module_score_gsva_filtered_samples_genes_df.Mono.BresciaLE.csv"))

# module.scores.df <- read.csv(paste(OUT_DIR, "module_score_gsva_filtered_samples_genes_df.Mono.BresciaLE.csv", sep = ""),
#                              row.names = 1)

# calculate mean for some duplicated samples
meta.sample <- meta[,c("sample_id","alt.subject.id", "visit", "sex", "race", "group", "flu.vax.count.10yr", "vaccine.side.effects", "long.covid.symptoms", "visit.overall.group.sex")] %>%
  distinct() %>%
  filter(alt.subject.id != "Control")

module.scores.df.sample <- module.scores.df %>% 
  group_by(celltype, geneset, sample_id) %>%
  dplyr::summarise(module.score.mean = mean(module.score)) %>%
  inner_join(meta.sample, by=c("sample_id"))

module.scores.df.sample <- module.scores.df.sample %>% 
  mutate(sex.group = factor(paste(sex, group, sep = "."), levels = c("Female.HC","Female.COVR","Male.HC","Male.COVR"))) %>%
  mutate(group.timepoint = factor(paste(group, visit, sep = "."),levels = c("HC.Day 0","HC.Day 1","HC.Day 28","COVR.Day 0","COVR.Day 1","COVR.Day 28"))) %>%
  mutate(group = factor(group, levels = c("HC","COVR")) )

module.scores.df.f <- filter(module.scores.df.sample, sex == "Female")
module.scores.df.m <- filter(module.scores.df.sample, sex == "Male")
```


p-values calculation for D1 vs D0, D28 vs D0 scores
Post-vaccine Model: use dream model to account for confounding factors and treat subject as random effect
```{r vaccine.response}
module.scores <- read.csv(file.path(OUT_DIR,"module_score_gsva_filtered_samples_genes_df.Mono.BresciaLE.csv"), row.names = 1)
genesets <- unique(module.scores$geneset)

model.res <- data.frame()
for (i in unique(module.scores$celltype)) {
  tryCatch({
    cat(i,"-\n")
    cell.type.modules <- subset(module.scores,celltype == i & !is.na(sex))
    cell.type.modules <- reshape2::dcast(cell.type.modules,
                                         Batch + alt.subject.id + visit + age + sex + race + group + flu.vax.count.10yr ~ geneset,value.var = "module.score")

    # using D0 only as pre-vax baseline
   exprObj = data.frame(t(cell.type.modules[,genesets]))
   colnames(exprObj) <- NULL
    timepoint.model.fit <- timepoint.condition.model(exprObj,cell.type.modules,c("Day 0"))
    
    # timepoint group diff
    timepoint.model.de.res <- model.term.statistics(timepoint.model.fit,grep("(D1|D28).v.baseline.COVR",
                                                     colnames(timepoint.model.fit$coefficients),value = T))
    model.res <- rbind(model.res,cbind(cell.type=i,timepoint.model.de.res))
  },error=function(err_msg){message(err_msg)})
}
```



### Boxplots for all timepoints in COVR group
**Extended Data Fig 5d and 5e**
```{r boxplot1, fig.width=20, fig.height=6}
# add pvalues from the dream model
model.res.p <- model.res %>% 
  separate(col = "term", into = c("group2","group"), sep = "\\.v\\.") %>%
  separate(col = "group", into = c("group1","group","sex"), sep = "\\.") %>%
  mutate(group1 = "Day 0") %>%
  mutate(group2 = sub("D", "Day ", x = group2)) %>%
  mutate(geneset = parameter)


# Female
celltypes <- unique(module.scores.df.f$celltype)
for (i in celltypes){
  df <- filter(module.scores.df.f, celltype == i)
  pval <- filter(model.res.p, cell.type == i, sex == "Female") %>% 
    arrange(parameter, group2) %>%
    mutate(y.position = rep(c(0.5,0.7),4)) %>%
    mutate(P.Value = ifelse(P.Value < 0.01, formatC(P.Value, format = "e", digits = 2), formatC(P.Value, format = "f", digits = 3)))
  p <- plotcelltype_group_covid(df) + ggtitle(i)
  print(p)
}

# Male
celltypes <- unique(module.scores.df.m$celltype)
for (i in celltypes){
  df <- filter(module.scores.df.m, celltype == i)
    pval <- filter(model.res.p, cell.type == i, sex == "Male") %>% 
    arrange(parameter, group2) %>%
    mutate(y.position = rep(c(0.5,0.7),4)) %>%
    mutate(P.Value = ifelse(P.Value < 0.01, formatC(P.Value, format = "e", digits = 2), formatC(P.Value, format = "f", digits = 3)))
  p <- plotcelltype_group_covid(df) + ggtitle(i)
  print(p)
}

```


### Boxplots for Day 0: COVR vs HC
**Supplementary Fig 2e**
```{r boxplot2, fig.width=20, fig.height=6}
module.scores.df.sample.D0 <- filter(module.scores.df.sample, visit == "Day 0")
module.scores.df.D0.f <- filter(module.scores.df.sample.D0, sex == "Female")
module.scores.df.D0.m <- filter(module.scores.df.sample.D0, sex == "Male")

# pdf("figures/module.score.Mono.BresciaLE.D0.SupFig2e.pdf", width = 20, height = 4.5)
# Female
df <- filter(module.scores.df.D0.f, celltype == "Mono_Classical")
p <- plotcelltype_group(df) + ggtitle(i)
print(p)

# Male
df <- filter(module.scores.df.D0.m, celltype == "Mono_Classical")
p <- plotcelltype_group(df) + ggtitle(i)
print(p)

# dev.off()
```

```{r}
sI <- sessionInfo()
utils:::print.sessionInfo(sI[-c(10,11)])
```

