---
title: "D0_scdotplot_CD8"
output: html_document
date: "2023-01-01"
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)  
library(Seurat)
library(tidyverse)
library(matrixStats)
library(parallelDist)
library(RColorBrewer)
```

## Generate the single cell dotplot for D0 COVR vs. HC Leading-edge genes for CD8 centeral memory and CD8 effector memory  
Download the Seurat object first and put it into input folder and read-in using the commented codes  
**Generate Figure 1h-1i dotplots**  

```{r input_seurat}
merge <- readRDS(file.path("data","covid_flu.CITEseq.Seurat.obj.RDS"))
merge$Timepoint.sex.group <- paste(merge$visit, merge$sex, merge$group,
                                 sep = ".")
CD8_Mem <- subset(merge, subset = cell.type %in% c("CD8_CM","CD8_EM"))
CD8_Mem <- subset(CD8_Mem, subset = visit %in% c("Day 0"))
# do a scTransform to remove batch
CD8_Mem <- SCTransform(CD8_Mem, assay = "RNA", method = "glmGamPoi", vars.to.regress = "Batch", verbose = FALSE)
DefaultAssay(CD8_Mem) <- "SCT"
```

format the LE sets to extract BTM-M7.3 T cell activation module LE from CD8-CM and CD8-EM  

```{r input_LEgenes}
selected.reset.M7.3.CM <- read.csv("data/LE_sets/pseudobulk.CD8_CM.vaccine.response.gsea.csv") %>%
  filter(term %in% c("Baseline.covid.f","Baseline.covid.m")) %>%
  filter(parameter %in% c("T cell activation (II) (M7.3)"))
selected.reset.M7.3.EM <- read.csv("data/LE_sets/pseudobulk.CD8_EM.vaccine.response.gsea.csv") %>%
  filter(term %in% c("Baseline.covid.f","Baseline.covid.m")) %>%
  filter(parameter %in% c("T cell activation (II) (M7.3)"))
CM.LE <- sapply(selected.reset.M7.3.CM$core_enrichment, function(x)str_split(x, "/"))
names(CM.LE) <- c("Female", "Male")
CM.LE <- c(CM.LE, list("m.and.f.union" = union(CM.LE[[1]], CM.LE[[2]]),
              "m.and.f.intersect" = intersect(CM.LE[[1]], CM.LE[[2]])))
EM.LE <- sapply(selected.reset.M7.3.EM$core_enrichment, function(x)str_split(x, "/"))
names(EM.LE) <- c("Male", "Female")
EM.LE <- c(EM.LE, list("m.and.f.union" = union(EM.LE[[1]], EM.LE[[2]]),
              "m.and.f.intersect" = intersect(EM.LE[[1]], EM.LE[[2]])))

selected.reset.M7.3 <- list("CD8_CM" = CM.LE,
                            "CD8_EM" = EM.LE)
# saveRDS(selected.reset.M7.3,"input/LE_sets/CD8_CM_EM.LE.M7.3activation.RDS")
```

select interested LE genes to show (limit the numbers of genes showed in figure to save space)  
showing example LE genes  
```{r norm_and_select_genes1}
CD8_Mem$sex.group <- factor(paste(CD8_Mem$sex, CD8_Mem$group, sep = "."), levels = c("Male.COVR","Male.HC","Female.COVR","Female.HC"))

# CD8-CM
CD8_CM <- subset(CD8_Mem, subset = cell.type == "CD8_CM")
Idents(CD8_CM) <- "group"
markers <- FindMarkers(CD8_CM, ident.1 = "COVR", ident.2 = "HC", min.pct = 0.1, only.pos = TRUE, logfc.threshold = 0)
# remove low expressed genes (<10%)
selected.genes.cm <- intersect(selected.reset.M7.3$CD8_CM$m.and.f.union, rownames(markers))

# CD8-EM
CD8_EM <- subset(CD8_Mem, subset = cell.type == "CD8_EM")
Idents(CD8_EM) <- "group"
markers <- FindMarkers(CD8_EM, ident.1 = "COVR", ident.2 = "HC", min.pct = 0.1, only.pos = TRUE, logfc.threshold = 0)
# remove low expressed genes (<10%)
selected.genes.em <- intersect(selected.reset.M7.3$CD8_EM$m.and.f.union, rownames(markers))
```


Dotplots  
**Fig 1h,1i**  
```{r dotplots, fig.width = 7, fig.height = 3.8}
# CD8-CM
# pdf(file = "figures/CD8_CM.Tactivation.baselineLE.dotplot.fig1h.pdf", width = 7.5, height = 3.5)
DotPlot(CD8_CM, features = selected.genes.cm, group.by = "sex.group", dot.scale = 10, cols = c("gray95", "blue")) + 
  RotatedAxis()+ggtitle("D0 (Baseline) LE genes")+theme(axis.text=element_text(size=16))+
  scale_size(range=c(0,10))
# dev.off()

# CD8-EM
# pdf(file = "figures/CD8_EM.Tactivation.baselineLE.dotplot.fig1i.pdf", width = 7.5, height = 3.5)
DotPlot(CD8_EM, features = selected.genes.em, group.by = "sex.group", dot.scale = 10, cols = c("gray95", "blue")) + 
  RotatedAxis()+ggtitle("D0 (Baseline) LE genes")+theme(axis.text=element_text(size=16))+
  scale_size(range=c(0,10))
# dev.off()
```

  
```{r}
sI <- sessionInfo()
utils:::print.sessionInfo(sI[-c(10,11)])
```



