---
title: "D1vsD0_pbulk_IFNG_score"
output: html_document
date: "2023-01-03"
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(edgeR)
library(GSVA)
library(Biobase)
library(reshape2)
library(tidyverse)
library(ggpubr)
```

## Get GSVA score of the filtered pseudobulk objects for Fig 2e and Extended Data Fig 3b
using the HALLMARK_INTERFERON_GAMMA_RESPONSE geneset for module scores
input is output from filtered and normalized pseudobulk objects  

**Generate Fig 2e and Extended Data Fig 3b boxplots**  
Seurat object needs to be downloaded and put in data/ folder, pseudobulk lists generated using the pooling functions  

```{r input}
# Generate pseudobulk lists and set parameters for pbulk input/output
DGELISTS_IN_PATH <- "output/CITE-seq/covid_flu.CITEseq.pseudobulk.gene.expr.filtered.RDS"
OUT_DIR <- "output/pbulk/gsva/"
dir.create(OUT_DIR, recursive = TRUE)

pbulk_list <- readRDS(DGELISTS_IN_PATH)

# read in geneset list
geneset.list <- readRDS("data/LE_sets/kegg_go_btm_reactome_foointerferon.rds")
hallmark_ifng <- list("hallmark_ifng" = geneset.list$HALLMARK_INTERFERON_GAMMA_RESPONSE)

# list of expression data by cell type
eset_list <- lapply(pbulk_list, function(dge){
  mat <- dge$normalizedExpr
  meta <- dge$samples
  ExpressionSet(assayData = mat, phenoData = AnnotatedDataFrame(meta))
})
```
  
Calculate module scores
```{r cal_score}
scores_list <- lapply(1:length(eset_list), function(eset){
  # cat(names(eset_list)[eset],"\n")
  # celltype.gsea.res <- selected.reset.M7.3[[names(eset_list)[eset]]]$CITEseq.baseline.LE.Union.genesets
  # if (nrow(celltype.gsea.res) > 0) {
  module.scores <- gsva(expr = eset_list[[eset]], gset.idx.list = hallmark_ifng, method = "gsva", parallel.sz = 16, min.sz = 5, verbose = FALSE)
  return(cbind(reshape2::melt(exprs(module.scores)),celltype= names(eset_list)[eset]))
  # }
})


module.scores.df <- do.call("rbind",scores_list)
colnames(module.scores.df) <- c("geneset","sample","module.score","celltype")
# add meta data and transform the dataframe into esetlist
meta <- pbulk_list$Mono$samples[,c("alt.subject.id", "age", "Batch","visit","sex","race","group","flu.vax.count.10yr","vaccine.side.effects","long.covid.symptoms", "visit.overall.group.sex")] 
meta$sample <- rownames(meta)
meta$sample_id <- paste(meta$alt.subject.id, meta$visit.overall.group.sex, sep = ".")

# add other info
module.scores.df$geneset_celltype <- paste0(module.scores.df$geneset,"--",module.scores.df$celltype)
module.scores.df <- left_join(module.scores.df, 
                              meta,
                              by=c("sample"))
write.csv(module.scores.df, file.path(OUT_DIR, "module_score_gsva_filtered_samples_genes_df.hallmarkIFNg.all.csv"))

# module.scores.df <- read.csv(paste(OUT_DIR, "module_score_gsva_filtered_samples_genes_df.hallmarkIFNg.all.csv", sep = "/"), 
#                              row.names = 1)
module.scores.df$age <- NULL
module.scores.df$group.timepoint <- paste(module.scores.df$group, module.scores.df$visit, sep = ".")
module.scores.df$group.timepoint <- factor(module.scores.df$group.timepoint, levels = c("HC.Day 0","HC.Day 1","HC.Day 28","COVR.Day 0","COVR.Day 1","COVR.Day 28"))
module.scores.df$sample_id <- paste(module.scores.df$alt.subject.id, module.scores.df$visit.overall.group.sex, sep = ".")

# calculate mean for some duplicated samples
meta.sample <- meta[,c("sample_id","alt.subject.id", "visit", "sex", "race", "group", "flu.vax.count.10yr", "vaccine.side.effects", "long.covid.symptoms", "visit.overall.group.sex")] %>%
  distinct() %>%
  filter(alt.subject.id != "Control")
module.scores.df.sample <- module.scores.df %>% group_by(celltype, geneset, sample_id) %>%
  dplyr::summarise(module.score.mean = mean(module.score)) %>%
  inner_join(meta.sample, by=c("sample_id"))

celltypes = c("CD4","Mono","B","cDC")

# calculate D1-D0, D28-D0 IFNG response score change
module.scores.df.change <- reshape2::dcast(module.scores.df.sample, celltype+geneset+alt.subject.id+sex+race+group+flu.vax.count.10yr+vaccine.side.effects+long.covid.symptoms ~ visit, value.var = "module.score.mean", fun.aggregate = mean) %>%
  mutate(D1vsD0 = `Day 1` - `Day 0`,
         D28vsD0 = `Day 28` - `Day 0`) %>%
  filter(celltype %in% celltypes) %>%
  mutate(group = factor(group, levels = c("HC","COVR")) ) %>%
  mutate(sex.group = factor(paste(sex, group, sep = "."), levels = c("Female.HC","Female.COVR","Male.HC","Male.COVR")))

module.scores.df.change$celltype <- factor(module.scores.df.change$celltype, levels = celltypes)
```
  
  
Boxplots of module scores  
**Fig 2e**  
```{r boxplot1, fig.width=8.5, fig.height=6.5}
my_comparisons1 <- list( c("Female.HC", "Female.COVR"), c("Female.HC", "Male.HC"), c("Female.COVR","Male.COVR"), c("Male.HC", "Male.COVR"))

# D1 vs D0
p1 <- ggplot(module.scores.df.change, aes(x = sex.group, y = D1vsD0)) +
  geom_boxplot(aes(fill=sex,color=sex), alpha = 0.4, width = 0.3, outlier.shape = NA)+
  geom_point(aes(fill=sex), shape = 21, color = "white", size = 3, position = position_jitterdodge(dodge.width = 0.9, jitter.width = 0.2)) +
  scale_color_manual(values = c("Female"="#BD3B29", "Male"="#0472B6"))+
  scale_fill_manual(values = c("Female"="#BD3B29", "Male"="#0472B6"))+
  stat_compare_means(comparisons = my_comparisons1, method = "wilcox.test", size = 5, label.y = c(1,1.25,1.5,1.75))+
  facet_wrap(~celltype, ncol = 2) +
  theme_bw(base_size = 20)+
  ylim(c(-0.3,2))+
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p1
```


**Extended Data Fig 3b**  
```{r boxplot2, fig.width=8.5, fig.height=6.5}
module.scores.df.sample.D0 <- filter(module.scores.df.sample, visit == "Day 0") %>%
  filter(celltype %in% celltypes) %>%
  mutate(celltype = factor(celltype, levels = celltypes)) %>%
  mutate(group = factor(group, levels = c("HC","COVR")) ) %>%
  mutate(sex.group = factor(paste(sex, group, sep = "."), levels = c("Female.HC","Female.COVR","Male.HC","Male.COVR")))

my_comparisons1 <- list( c("Female.HC", "Female.COVR"), c("Female.HC", "Male.HC"), c("Female.COVR","Male.COVR"), c("Male.HC", "Male.COVR"))

# plot only D0 (baseline)
p2 <- ggplot(module.scores.df.sample.D0, aes(x = sex.group, y = module.score.mean)) +
  geom_boxplot(aes(fill=sex,color=sex), alpha = 0.4, width = 0.3, outlier.shape = NA)+
  geom_point(aes(fill=sex), shape = 21, color = "white", size = 3, position = position_jitterdodge(dodge.width = 0.9, jitter.width = 0.2)) +
  scale_color_manual(values = c("Female"="#BD3B29", "Male"="#0472B6"))+
  scale_fill_manual(values = c("Female"="#BD3B29", "Male"="#0472B6"))+
  stat_compare_means(comparisons = my_comparisons1, method = "wilcox.test", size = 5, label.y = c(0.5,0.75,1,1.25))+
  facet_wrap(~celltype, ncol = 2) +
  theme_bw(base_size = 20)+
  ylim(c(-0.6,1.4))+
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
p2
```




```{r}
sI <- sessionInfo()
utils:::print.sessionInfo(sI[-c(10,11)])
```



