---
title: "D0D1_scdotplot_AgPre_Mono"
output: html_document
date: "2023-01-04"
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(Seurat)
library(tidyverse)
library(matrixStats)
library(parallelDist)
library(RColorBrewer)
```

## Generate the single cell dotplot plot for Antigen presentation genesets Leading-edge genes for classical monocytes  
Download the Seurat object first and put it into data folder and read-in using the commented codes  
Leading edge genes from dream model, D1.covid.m.sub.baseline term, indicating D1 COVR-M change differences after accounting for baseline COVR-M vs. HC-M  
Antigen presentation genesets:  
BTM-enriched in antigen presentation (I) (M71)  
BTM-enriched in antigen presentation (II) (M95.0)  

**Generate Figure 2f dotplots**

```{r input_seurat}
merge <- readRDS(file.path("data","covid_flu.CITEseq.Seurat.obj.RDS"))
merge$visit.sex.group <- paste(merge$visit, merge$sex, merge$group,
                                   sep = ".")
monoc <- subset(merge, subset = visit %in% c("Day 0","Day 1") & cell.type == "Mono_Classical")
DefaultAssay(monoc) <- "SCT"

### input LE sets
Agpre.LE <- readRDS("data/LE_sets/mono.Agpresentation.D1.covid.m.sub.baseline.LE.rds")
```


dotplot for Antigen presentation LE
Classical monocytes
```{r dotplots, fig.width = 8, fig.height = 5.5}
monoc$visit.sex.group <- factor(monoc$visit.sex.group, 
                         levels = c("Day 1.Male.COVR", "Day 0.Male.COVR",
                                    "Day 1.Male.HC", "Day 0.Male.HC",
                                    "Day 1.Female.COVR","Day 0.Female.COVR",
                                    "Day 1.Female.HC", "Day 0.Female.HC"))
# put HLA genes first
HLA_sets <- Agpre.LE[str_detect(Agpre.LE, pattern = "HLA")]
Agpre.LE <- unique(c(HLA_sets, Agpre.LE))

# pdf(file = "figures/monoc.Agpre.D1xgroupmaleLE.dotplot.fig2f.pdf", width = 8, height = 5.5)
DotPlot(monoc, features = Agpre.LE, group.by = "visit.sex.group", dot.scale = 10, cols = c("gray95", "blue"))+
  RotatedAxis()+theme(axis.text=element_text(size=12))+
  scale_size(range=c(0,10))
# dev.off()
```


```{r}
sI <- sessionInfo()
utils:::print.sessionInfo(sI[-c(10,11)])
```





