---
title: "COVID Flu: CITE-Seq Innate Immune Receptor (IIR) Signature"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: True
editor_options:
  chunk_output_type: console
---

IIR signature is the combination of BTM M11.0 and M4.0 genesets

#### Setup
```{r setup,message=FALSE}
library(ggplot2)
library(knitr)
library(dplyr)
source('R/utils/visualization.R')
source('R/utils/dream.models.R')
loadLibraries()

knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE, out.width=900)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE, out.width=900)

output.folder <- file.path("output","CITE-seq")
dir.create(output.folder)
figure.folder <- "figures"
dir.create(figure.folder)

# cite-seq pbulk data
pseudobulk.list.filtered <- readRDS(file.path(output.folder,"covid_flu.CITEseq.pseudobulk.gene.expr.filtered.RDS"))
cat("Cell subsets:",names(pseudobulk.list.filtered))

# genesets and celltypes of interest
wb.genesets <- c("enriched in monocytes (II) (M11.0)","cell cycle and transcription (M4.0)")
mono.subsets <- c("Mono_Classical","Mono_NonClassical")

dea.files <- list.files(file.path(output.folder,"DEA"),pattern = "vaccine.response.gene",full.names = T)
cite_de_model_results <- do.call(rbind,lapply(dea.files[grep(paste0(mono.subsets,collapse = "|"),dea.files)],read.csv))
gsea.files <- list.files(file.path(output.folder,"GSEA"),pattern = "vaccine.response.gsea",full.names = T)
gsea.results <- do.call(rbind,lapply(gsea.files[grep(paste0(mono.subsets,collapse = "|"),gsea.files)],read.csv))
```

#### Baseline leading-edge genes
LE genes from IIR signature in mono subsets
```{r reset}
baseline.IIR.geneset <- subset(gsea.results,parameter %in% wb.genesets & term %in% c("Baseline.COVR.f","Baseline.COVR.m") & 
                                   cell.type %in% mono.subsets)
baseline.IIR.geneset$core_enrichment <- sapply(baseline.IIR.geneset$core_enrichment,strsplit,"/",fixed=T)
baseline.IIR.geneset <- reshape2::dcast(baseline.IIR.geneset,cell.type + parameter ~ term,value.var = "core_enrichment")

# merge all LE genes within a cell type
for (i in unlist(unique(baseline.IIR.geneset$cell.type))) {
  tmp <- subset(baseline.IIR.geneset,cell.type == i)
  union.genes <- lapply(3:4,function(x){unique(unlist(tmp[,x]))})
  union.genes <- cbind(cell.type = i,parameter = "Union.genesets",Baseline.COVR.f=union.genes[1],Baseline.COVR.m=union.genes[2])
  baseline.IIR.geneset <- rbind(baseline.IIR.geneset,union.genes)
}
colnames(baseline.IIR.geneset)[3:4] <- c("Female","Male")
```

#### Post-vaccination shift of genes in baseline genesets
Identify LE genes that moved towards baseline HC after vaccination
```{r genesets}
# COVR vs. HC
overall.covid.v.baseline.hc <- subset(cite_de_model_results,contrast %in% c("COVR.v.baseline.HC.f","COVR.v.baseline.HC.m"))
overall.covid.v.baseline.hc <- rbind(overall.covid.v.baseline.hc,subset(cite_de_model_results,term %in% c("Baseline.COVR.f","Baseline.COVR.m")))
overall.covid.v.baseline.hc$sex <- "Female"
overall.covid.v.baseline.hc[grep(".m$",overall.covid.v.baseline.hc$term),"sex"] <- "Male"

overall.cell.type.genesets <- list()
overall.covid.v.baseline.hc.new <- data.frame()
for (i in mono.subsets) {
  cat(i,"-\n")
  cell.type.genesets <- list()
  covid.v.baseline.hc <- subset(overall.covid.v.baseline.hc,cell.type == i)
  
  # resolving is defined as smaller abs z.std value at D1 or D28 compared to D0
  post.vax.resolving.genes <- as.data.frame(covid.v.baseline.hc %>% group_by(sex,gene.symbol) %>% 
  summarize(D1.resolving = (abs(z.std[visit == "D1"]) < abs(z.std[visit == "Baseline"])),
             D28.resolving = (abs(z.std[visit == "D28"]) < abs(z.std[visit == "Baseline"]))))
  covid.v.baseline.hc <- merge(covid.v.baseline.hc,post.vax.resolving.genes,by=c("sex","gene.symbol"))
  overall.covid.v.baseline.hc.new <- rbind(overall.covid.v.baseline.hc.new,covid.v.baseline.hc)
  
  # cell type specific D0 covid vs. hc DE genes
  baseline.de.genes <- subset(covid.v.baseline.hc,P.Value < 0.05 & visit == "Baseline" & contrast %in% c("covid.m","covid.f"))
  baseline.de.genes <- baseline.de.genes %>% summarise(geneset="Baseline DE genes",
             male_genes = list(gene.symbol[sex == "Male"]),
             female_genes = list(gene.symbol[sex == "Female"]))
  # add baseline DE genes to gene list
  cell.type.genesets$CITEseq.baseline.DE.genes["Male"] <- baseline.de.genes$male_genes
  cell.type.genesets$CITEseq.baseline.DE.genes["Female"] <- baseline.de.genes$female_genes

  # D1 reset genes, not necessarily DE at baseline
  reset.genes <- subset(covid.v.baseline.hc,sex == "Female" & D1.resolving == T)
  cell.type.genesets$CITEseq.D1.reset.genes[["Female"]] <- unique(reset.genes$gene.symbol)
  reset.genes <- subset(covid.v.baseline.hc,sex == "Male" & D1.resolving == T)
  cell.type.genesets$CITEseq.D1.reset.genes[["Male"]] <- unique(reset.genes$gene.symbol)
  
  # D1 and D28 reset genes, not necessarily DE at baseline
  reset.genes <- subset(covid.v.baseline.hc,sex == "Female" & D1.resolving == T & D28.resolving == T)
  cell.type.genesets$CITEseq.stable.reset.genes[["Female"]] <- unique(reset.genes$gene.symbol)
  reset.genes <- subset(covid.v.baseline.hc,sex == "Male" & D1.resolving == T & D28.resolving == T)
  cell.type.genesets$CITEseq.stable.reset.genes[["Male"]] <- unique(reset.genes$gene.symbol)
  
  # baseline DE genes that are reset at D1
  reset.genes <- subset(covid.v.baseline.hc,sex == "Female" & D1.resolving == T & gene.symbol %in% unique(unlist(baseline.de.genes$female_genes)))
  cell.type.genesets$CITEseq.baseline.DE.D1.reset.genes[["Female"]] <- unique(reset.genes$gene.symbol)
  # baseline DE genes that are reset at D1
  reset.genes <- subset(covid.v.baseline.hc,sex == "Male" & D1.resolving == T & gene.symbol %in% unique(unlist(baseline.de.genes$male_genes)))
  cell.type.genesets$CITEseq.baseline.DE.D1.reset.genes[["Male"]] <- unique(reset.genes$gene.symbol)
  
  # baseline DE genes that are reset at D1 and D28
  reset.genes <- subset(covid.v.baseline.hc,sex == "Female" & D1.resolving == T & D28.resolving == T &
                          gene.symbol %in% unique(unlist(baseline.de.genes$female_genes)))
  cell.type.genesets$CITEseq.baseline.DE.stable.reset.genes[["Female"]] <- unique(reset.genes$gene.symbol)
  # baseline DE genes that are reset at D1
  reset.genes <- subset(covid.v.baseline.hc,sex == "Male" & D1.resolving == T & D28.resolving == T &
                          gene.symbol %in% unique(unlist(baseline.de.genes$male_genes)))
  cell.type.genesets$CITEseq.baseline.DE.stable.reset.genes[["Male"]] <- unique(reset.genes$gene.symbol)
  
  
  # intersect above genes with genesets of interest
  tmp <- subset(baseline.IIR.geneset,cell.type == i)
  expr.status <- names(cell.type.genesets)
  for (g in 1:nrow(tmp)) {
    for (s in c("Male","Female")) {
      # LE genes
      cell.type.genesets[[paste0("CITEseq.baseline.LE.",make.names(tmp[g,"parameter"]))]][s] <- tmp[g,s]
      for (ex in expr.status) {
        cell.type.genesets[[paste0(ex,".LE.",make.names(tmp[g,"parameter"]))]][s] <- 
          list(intersect(unlist(tmp[g,s]), cell.type.genesets[[ex]][[s]]))
      }
    }
  }
  # union and intersect of m and f genesets
  for (g in names(cell.type.genesets)) {
    cell.type.genesets[[g]][["m.and.f.union"]] <- union(cell.type.genesets[[g]]$Female,cell.type.genesets[[g]]$Male)
    cell.type.genesets[[g]][["m.and.f.intersect"]] <- intersect(cell.type.genesets[[g]]$Female,cell.type.genesets[[g]]$Male)
  }
  overall.cell.type.genesets[[i]] <- cell.type.genesets
}
# union of all cell subsets
genesets.to.merge <- names(overall.cell.type.genesets$Mono_Classical)
for (g in genesets.to.merge) {
  tmp <- lapply(mono.subsets,function(x){overall.cell.type.genesets[[x]][[g]]})
  names(tmp) <- mono.subsets
  for (s in names(tmp$Mono_Classical)) {
    genes <- lapply(mono.subsets,function(x){tmp[[x]][[s]]})
    names(genes) <- mono.subsets
    if (length(unlist(genes)) > 0) {
      #gplots::venn(genes)
      #title(paste0(g," - ",s))
      overall.cell.type.genesets[["Monocytes"]][[g]][[s]] <- unique(unlist(genes))
    }
  }
}
saveRDS(overall.cell.type.genesets,file.path(output.folder,"mono.LE.and.reset.genesets.RDS"))
geneset.name <- "CITEseq.baseline.LE.Union.genesets" # IIR signature
```

#### Geneset-level shifts
Fig. 4b and c: Use all LE genes of the union geneset to create modules and see how they change at different timepoints
```{r modules,fig.cap=c("Fig. 4b","Fig. 4c")}
pdf(file.path(figure.folder,paste0("Fig4bc.pdf")),width=6,height=4.5)

plots <- list()
for (i in mono.subsets) {
  normalized.expr.batch.effect.filtered <- pseudobulk.list.filtered[[i]]$normalizedExpr
  geneset.list <- overall.cell.type.genesets[[i]][[geneset.name]]
  sample.signature.scores <- GSVA::gsva(normalized.expr.batch.effect.filtered,
                                          gset.idx.list = geneset.list,
                                          parallel.sz=1)
    
  sample.info <- pseudobulk.list.filtered[[i]]$samples
  sample.info <- cbind(sample.info,t(sample.signature.scores)[rownames(sample.info),])
  # combine the sex-specific scores accordingly
  sample.info$module.score <- ifelse(sample.info$sex == "Male",sample.info$Male,sample.info$Female)

  # model of timepoint differences
  sample.info <- subset(sample.info,!is.na(visit))
  timepoint.model.fit <- timepoint.condition.model(t(sample.info[,c("module.score","m.and.f.union")]),sample.info,c("Day 0"))
  model.de.res <- model.term.statistics(timepoint.model.fit,grep("[0-9].v.baseline.COVR",colnames(timepoint.model.fit$coefficients),value=T))
  print(subset(model.de.res,parameter == "module.score"))
  model.de.res <- model.term.statistics(timepoint.model.fit,grep("[0-9].v.baseline.HC",colnames(timepoint.model.fit$coefficients),value=T))
  print(subset(model.de.res,parameter == "module.score"))

  sample.info.mean <- as.data.frame(sample.info %>% 
    group_by(alt.subject.id,visit,group,sex,age,race,flu.vax.count.10yr,covid.diagnosis.start.date.to.sample.drawn) %>% 
    summarise(module.score=mean(module.score),union.score=mean(m.and.f.union)))
  sample.info.mean$group <- factor(sample.info.mean$group,levels = c("HC","COVR"))
  
  # median of baseline HC
  hc.baseline.median <- as.data.frame(subset(sample.info.mean,visit == "Day 0" & group == "HC") %>% 
                  group_by(sex) %>% summarise(median.score=median(module.score)))

  plots[[i]] <- ggplot(sample.info.mean, aes(x = visit, y = module.score)) +
    geom_boxplot(aes(fill=sex),color = "black", alpha = 0.4, width = 0.75, outlier.shape = NA) +
    geom_line(aes(group=alt.subject.id,color=sex),alpha=0.75) +
    geom_point(aes(color=sex),size=0.5) +
    scale_color_manual(values = c("Female"="#BD3B29", "Male"="#0472B6"))+
    scale_fill_manual(values = c("Female"="#BD3B29", "Male"="#0472B6"))+
    stat_compare_means(comparisons = list(c(1,2),c(1,3)), method = "wilcox.test", size = 3, paired=T) +
    facet_grid(sex~group) +
    scale_y_continuous(expand = c(0.03,0.17)) +
    theme_bw(base_size = 14) + ggtitle(i) + ylab("Module score") + xlab("Timepoint") +
    geom_hline(data=hc.baseline.median,aes(yintercept=median.score,color=sex),linetype="dashed",size=1,alpha=0.4)
  print(plots[[i]])
}
dev.off()
plots
```

#### Gene-level shifts
ED Fig. 5a and b
```{r ed.fig.5ab,fig.cap=c("ED Fig. 5a","ED Fig. 5b")}
# visualize for each gene set
pdf(file.path(figure.folder,"ED.Fig.5ab.pdf"),width=6,height=4)
to.highlight=list(Mono_Classical=c("NOD2","TLR2","CLEC4E"),Mono_NonClassical=c("TLR4","FPR2","ITGAM"))
visit.colors <- c(Baseline="#CDF0EA",D1="grey80",D28="#C7D36F")

plots <- list()
for (i in mono.subsets) {
  cat(i,"-\n")
  cell.type.genesets <- overall.cell.type.genesets[[i]]
  covid.v.baseline.hc <- subset(overall.covid.v.baseline.hc.new,cell.type == i)
  
  # separate sex
  geneset.fc <- rbind(subset(covid.v.baseline.hc,sex == "Female" & gene.symbol %in% cell.type.genesets[[geneset.name]]$Female),
                      subset(covid.v.baseline.hc,sex == "Male" & gene.symbol %in% cell.type.genesets[[geneset.name]]$Male))
  
  # wilcoxon test
  cat("Wilcoxon test p value for post-vaccination gene expression changes:\n")
  print(reshape2::dcast(geneset.fc,gene.symbol + sex ~ visit,value.var = "z.std") %>% group_by(sex) %>%
    summarise(D1.pval=wilcox.test(D1,Baseline,paired=T)$p.value,D28.pval=wilcox.test(D28,Baseline,paired=T)$p.value))
  
  # reverse the order of the visits
  geneset.fc$visit <- factor(geneset.fc$visit,levels=c("D28","D1","Baseline"))
  if (nrow(geneset.fc) > 0) {
    plots[[i]] <- geneset.ridge.plot(geneset.fc,to.highlight[[i]]) + ggtitle(i) + scale_y_discrete(expand=c(0.05,0.05)) + 
      scale_fill_manual(values = visit.colors)
    print(plots[[i]])
  }
}
dev.off()
plots
```

#### Session Info
```{r session}
sessionInfo()
```

