---
title: "COVID Flu: CITE-Seq Innate Immune Receptor (IIR) Signature"
output:
  html_document:
    df_print: paged
    code_folding: hide
    toc: True
editor_options:
  chunk_output_type: console
---

Run a 20-round bootstrap test to determine whether the number of reset genes is significantly different COVR-F and COVR-M

#### Setup
```{r setup,message=FALSE}
library(ggplot2)
library(knitr)
library(dplyr)
library(BiocParallel)
library(doParallel)
param = SnowParam(32,"SOCK",progressbar = T)
register(param)
registerDoParallel()
options(future.globals.maxSize = 120 * 1024 ^3)
source('R/utils/visualization.R')
source('R/utils/dream.models.R')
loadLibraries()

knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
opts_knit$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE, out.width=900)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache.lazy = FALSE, out.width=900)

output.folder <- file.path("output","CITE-seq")
dir.create(output.folder)
figure.folder <- "figures"
dir.create(figure.folder)

# subject list
subject.IDs <- read.csv(file.path("data","cohort.subject.IDs.txt"),header = F)[,1]

# cite-seq pbulk data
pseudobulk.list.filtered <- readRDS(file.path(output.folder,"covid_flu.CITEseq.pseudobulk.gene.expr.filtered.RDS"))
cat("Cell subsets:",names(pseudobulk.list.filtered))

# genesets and celltypes of interest
wb.genesets <- c("enriched in monocytes (II) (M11.0)","cell cycle and transcription (M4.0)")
mono.subsets <- c("Mono_Classical","Mono_NonClassical")

dea.files <- list.files(file.path(output.folder,"DEA"),pattern = "vaccine.response.gene",full.names = T)
cite_de_model_results <- do.call(rbind,lapply(dea.files[grep(paste0(mono.subsets,collapse = "|"),dea.files)],read.csv))
gsea.files <- list.files(file.path(output.folder,"GSEA"),pattern = "vaccine.response.gsea",full.names = T)
gsea.results <- do.call(rbind,lapply(gsea.files[grep(paste0(mono.subsets,collapse = "|"),gsea.files)],read.csv))

# IIR Signature
overall.cell.type.genesets <- readRDS(file.path(output.folder,"mono.LE.and.reset.genesets.RDS"))
```

#### Monocyte IIR Signature
Genes from IIR signature in mono subsets
```{r IIR.sig}
mono.samples <- subset(pseudobulk.list.filtered$Mono_Classical$samples,!is.na(visit))
mono.samples$sample.id <- rownames(mono.samples)
# to ensure the subjects are ordered the same way as the original IDs used so the same set of samples would be drawn in each round of the bootstrapping
mono.samples$alt.subject.id <- factor(mono.samples$alt.subject.id,levels=subject.IDs) 
mono.samples.subjects <- unique(mono.samples[,c("alt.subject.id","group","sex")])
mono.samples.subjects$overall.group.sex <- factor(paste0(mono.samples.subjects$group,"-",mono.samples.subjects$sex),
                                                  levels = c("COVR-Female","HC-Female","COVR-Male","HC-Male"))

IIR.sig <- "CITEseq.baseline.LE.Union.genesets"

# prepare the variancePartition object for the bootstrapping
model <- as.formula(paste0("~ 0 + visit.overall.group.sex + age + race + flu.vax.count.10yr + n_barcodes + (1|Batch) + (1|alt.subject.id)"))
celltype.vobj <- list()
for (cell.type in c("Mono_Classical","Mono_NonClassical")) {
  samples <- pseudobulk.list.filtered[[cell.type]]
  include.samples <- samples$samples[]
  
  vobjDream <- voomWithDreamWeights(samples, model, include.samples,plot=T)
  celltype.vobj[[cell.type]] <- vobjDream
}
```

#### Bootstrapping
Bootstrap cohort of subjects

Reset genes are defined as those with smaller COVR vs baseline HC effect size at D1 and D28 compared to D0
```{r bootstrap}
num.rounds <- 20
overall.response.model.de.res <- data.frame()
for (i in 1:num.rounds) {
  cat("Round",i,"\n")
  set.seed(i)
  # select subjects from each group
  selected.subjects <- mono.samples.subjects %>% group_by(overall.group.sex) %>% slice_sample(prop=1,replace=T)
  # differentiate duplicates
  mono.samples.subset <- do.call(rbind,lapply(1:nrow(selected.subjects),
                                              function(x) {
                                                tmp <- subset(mono.samples,alt.subject.id == selected.subjects$alt.subject.id[x])
                                                tmp$index <- x
                                                return(tmp)}))
  # only 1 sample from replicates
  mono.samples.subset <- as.data.frame(mono.samples.subset %>% group_by(alt.subject.id,index,visit) %>% slice_sample(n=1))
  mono.samples.subset$alt.subject.id <- paste0(mono.samples.subset$alt.subject.id,".",mono.samples.subset$index)
  rownames(mono.samples.subset) <- make.unique(mono.samples.subset$sample.id)
  
  for (cell.type in names(celltype.vobj)) {
    tryCatch({
      cat(cell.type,"-\n")
      genes <- overall.cell.type.genesets[[cell.type]][[IIR.sig]][["m.and.f.union"]]  # union LEGs
      
      vobjDream <- celltype.vobj[[cell.type]]
      vobjDream.round <-subsetListOfArrays(vobjDream,j=mono.samples.subset$sample.id,i=genes,I=NA,IX=c(),IJ=c("E","weights"),JX=c("targets"))
      colnames(vobjDream.round$E) <- rownames(vobjDream.round$targets) 
      
      # using D0 only as pre-vax baseline
      timepoint.model.fit <- timepoint.condition.model(vobjDream.round,mono.samples.subset,c("Day 0"),
                                                       additional.random.effects = "Batch",additional.fixed.effects = "n_barcodes")
      # timepoint group diff
      timepoint.model.de.res <- model.term.statistics(timepoint.model.fit,
                                                      grep("Baseline.COVR.[mf]$|v.baseline.HC.[mf]$",
                                                           colnames(timepoint.model.fit$coefficients),value = T))
      
      overall.response.model.de.res <- rbind(overall.response.model.de.res,cbind(round=i,celltype=cell.type,timepoint.model.de.res))
      
    },error=function(err_msg){message(err_msg)})
  }
}
saveRDS(overall.response.model.de.res,file.path(output.folder,"mono.sig.bootstrapping.DE.res.RDS"))
```

#### Distribution of reversal genes
Fraction of genes that reset in each round

To have bootstrapped p-values exactly identical to the numbers reported in the paper, the published alt.subject.id need to be ordered  
the same way as the internal IDs used when generating the figure, so that the same samples are drawn in each round of bootstrapping.
```{r reset,fig.cap="Fig. 4e"}
overall.response.model.de.res[,c("visit","contrast")] <-  
        do.call(rbind,sapply(overall.response.model.de.res$term,stringi::stri_split_fixed,pattern=".",n=2))
overall.response.model.de.res$sex <- "Female"
overall.response.model.de.res[grep(".m$",overall.response.model.de.res$contrast),"sex"] <- "Male"

reset.stat <- data.frame()
for (s in unique(overall.response.model.de.res$sex)) {
  for (cell.type in unique(overall.response.model.de.res$celltype)) {
    res <- subset(overall.response.model.de.res,sex == s & celltype == cell.type)
    res <- reshape2::dcast(res,round + parameter ~ visit,value.var = "z.std")
    res$D1.reset <- abs(res$Baseline) > abs(res$D1)
    res$D28.reset <- abs(res$Baseline) > abs(res$D28)
    res$reset <- res$D1.reset & res$D28.reset
    # only sex-specific LEGs
    mono.sig.LE.genes <- overall.cell.type.genesets[[cell.type]][[IIR.sig]][[s]]
    res <- subset(res,parameter %in% mono.sig.LE.genes)
    
    num.reset <- as.data.frame(res %>% group_by(round) %>% summarise(reset.count=sum(reset)))
    num.reset$proportion.reset <- num.reset$reset.count/length(mono.sig.LE.genes)
    reset.stat <- rbind(reset.stat,cbind(celltype=cell.type,sex=s,num.LEG=length(mono.sig.LE.genes),num.reset))
  }
}

# calculate mean, 95% confidence intervals, and wilcox test p values
reset.stat.summary <- data.frame()
wilcox.test.res <- data.frame()
for (i in unique(reset.stat$celltype)) {
  test.res <- wilcox.test(subset(reset.stat,celltype == i & sex == "Male")$proportion.reset,
                          subset(reset.stat,celltype == i & sex == "Female")$proportion.reset)
  wilcox.test.res <- rbind(wilcox.test.res,cbind(celltype=i,p=sprintf("%.1e",test.res$p.value),"group1"="Male","group2"="Female"))
  for (j in unique(reset.stat$sex)) {
    l.model <- lm(proportion.reset ~ 1,subset(reset.stat,celltype == i & sex == j))
    model.summary <- summary(l.model)
    model.res <- as.data.frame(confint(l.model,level=0.95))
    model.res$mean <- model.summary$coefficients[,"Estimate"]
    reset.stat.summary <- rbind(reset.stat.summary,cbind(celltype=i,sex=j,model.res))
  }
}
colnames(reset.stat.summary)[3:4] <- c("low","high")

ggplot(reset.stat.summary, aes(x = sex, y = mean)) +
    geom_point(aes(color=sex)) + geom_errorbar(aes(ymin=low,ymax=high,color=sex),width=.3) +
    scale_color_manual(values = c("Female"="#BD3B29", "Male"="#0472B6"))+
    scale_fill_manual(values = c("Female"="#BD3B29", "Male"="#0472B6"))+
    stat_pvalue_manual(wilcox.test.res,y.position = 0.5) +
    facet_grid(~celltype) + scale_y_continuous(limits = c(0.15,0.7)) +
    theme_light(base_size = 11) + ylab("Proportion of LEGs moved towards baseline HC") + xlab("") +
  scale_x_discrete(limits=rev) + coord_flip() 
ggsave(file.path(figure.folder,"Fig4e.pdf"),width=5,height=2)
```

#### Session Info
```{r session}
sessionInfo()
```

